---
title: "Heart Failure Cohort Transcription Factor Enchriment"
author: "Aaron Troy"
date: "2023-05-01"
output:
  pdf_document: default
  html_document: default
---

#### Transcription Factor Enrichment for Individual Patients Using VIPER and prior knowledge of transcription factor regulatory relationships 

The purpose of this notebook is to leverage prior knowledge on transcription factor --> target gene relatioships to rank which TFs are most likley implicated in observed expression profiles. Here, this is done in a groupwise fashion, rather than sample-by-sample. 


To do this, we require the following:

- Expression signatures associated with one or more conditions of interest. These can be simply normalized counts or the results of DEGs analysis, as in the case here. We will use the DESeq results with 3 paired contrasts between control, HFpEF, and HFrEF
- A prior knowledge network of TF-->gene relationships. These should be signed and included some confidence score. May databases exist for this. Here, we will use Dorothea (https://saezlab.github.io/dorothea/) with a threshold applied to select only high confidence interactions ]
- A statistical method for computing an enrichment score. For this, we use VIPER (https://bioconductor.org/packages/release/bioc/html/viper.html)

Start by importing the required libraries 
```{r setup}
knitr::opts_chunk$set(echo = TRUE)

options(warn=-1)

suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(tidyverse))
suppressMessages(library(biomaRt))
suppressMessages(library(modelr))
suppressMessages(library(poolr))
suppressMessages(library(pheatmap))
suppressMessages(library(lsa))
suppressMessages(library(DESeq2))
suppressMessages(library(GGally))

#Dorothea - transcription factor network
library(dorothea)

#VIPER - for inferring active TFs
library(viper)

#Human Protein Atlas
library(hpar)

#Metric to perform enrichment on. {'z_score', 'log2FoldChange', 'stat'}
metric = 'z_score'

```

## Retrieve Transcription Factor Network 

Here we select only links with A, B, or C grade confidence. See doRothea documentation to see how these categories are defined.  

Also take a peak at the distribution of regulon sizes after threshold. You can see most TFs have less than 100 targets 

```{r getDorothea}
dorotheaPKN <- dorothea::dorothea_hs %>% 
  dplyr::filter(confidence %in% c("A", "B", "C")) %>%
  dplyr::select(target, tf, mor) %>%
  as.data.frame()

n_genes <- dorotheaPKN %>%
  group_by(tf) %>%
  summarize(n = n())

ggplot(data=n_genes, aes(x=n)) +
  geom_density() +
  theme(text = element_text(size=12)) +
  xlab('Number of target genes') +
  ylab('densities') +
  theme_bw() +
  theme(legend.position = "none")

#VIPER reguires the PKN as a regulon object. Make this using dorothea
regulon <- df2regulon(dorotheaPKN)
```

## Filter TFs using the Human Protein Atlas 
```{r removeNonHeartTFs}
data(hpaNormalTissue)

# Querying the HPA requires ensembl IDs. Retrieve these for the TFs in the PKN using biomaRt
mart <- useEnsembl(biomart = 'genes', dataset = 'hsapiens_gene_ensembl', version = 92) 
map <- getBM(filters= "external_gene_name", attributes= c("ensembl_gene_id", "external_gene_name", "gene_biotype"),values=dorotheaPKN$tf, mart = mart, uniqueRows=T)  

# Filter for presense in the heart  
heartTFs <- getHpa(map$ensembl_gene_id) %>%
  filter(Tissue == "heart muscle") %>%
  filter(Level != "Not detected")
dorotheaPKN <- filter(dorotheaPKN, tf %in% heartTFs$Gene.name)

```

## Read in Data

Read the DE results from DESeq2  

```{r getDEGData}

setwd('~/Bioinformatics/Network Analysis of HFpEF/Hahn_FGN_Analysis/data')

# Sources for DEG and counts data for TF enrichement 
DEG_src = "cov_adj_DEGs_deseq.csv"
cnts_src = "cov_adj_normCounts_deseq.csv"

# DEGs
degs <- read.csv(DEG_src, check.names = FALSE, sep = ',', row.names = 1)

# Counts  
cnts <- read.csv(cnts_src, check.names = FALSE, sep = ',', row.names = 1)%>%
  as.tibble()%>%
  column_to_rownames('external_gene_name') %>%
  dplyr::select(!c('Row.names'))

#Phenotype data
pData <- read.csv("clinicalData.csv", check.names=FALSE) %>%
  filter(id %in% colnames(cnts)) %>%
  arrange(match(id, (colnames(cnts)))) %>%
  column_to_rownames("id")

head(pData)

#Assemble as expression set object for sample_by_sample analysis 
exp_dset <- ExpressionSet(assayData = as.matrix(cnts),
                          phenoData = AnnotatedDataFrame(pData))
                          

```
## msVIPER for TF Enrichment 

msVIPER works by computing a three tailed enrichment score. The routine can be summarized through the following steps:

- Genes are ranked by the absolute value of the input statistic. Many metrics are acceptable, including $z$-score, $\log_2$ fold change, or t-statistic. The resulting rankings are then quantile transformed 
- A one tail test is applied to compute the first enrichment score $ES_k^1$ for each regulator $k$:

$$
ES_k^1 = \frac{1}{N_k}\sum_{i = 1}^{N_k} (1 - |s_i|)l_iR(|x_i|)
$$
Here, $N_k$ is the number of target gene for regulator $k$, $s_i$ the mode of interaction (activation or inhibition), $l_i$ is the likelihood of the interaction, and $R(|x_i|)$ is the quantile transformed rank of the absolute value of statistic $x_i$. For us, this term always goes to zero, as $s_i \in \{-1,1\}$ for the dorothea network

- A second enrichment score $ES_k^2$ accounting for the mode of the interaction is then computed using:

$$
ES_k^2 = \frac{1}{N_k}\sum_{i = 1}^{N_k} l_iR(x_i)
$$
- Enrichment scores are then combined:

$$
ES_k = (ES_k^1 + |ES_k^2|) * \text{sign}(ES_k^2)
$$
- A final normalized enrichment score (NES) is then computed:

$$
NES_k = ES \sqrt{\sum_{i = 1}^{N_k} l_i^2}
$$
By default, VIPER assumes a uniform distribution of expression values as a prior. In some ways this is poor choice, as genes are heavily co-regulated in any circumstance and thus this choice will result in p-value under estimation. An alternative you might consider is permuting samples randomly, but in this case we only have 3 groups. 

TODO: think more about an appropriate prior. For the time being, let's use the null model based on permuting the gene rather than the samples. 

```{r msVIPER_DEGs}
#Build an expression matrix using results from DESeq
expMatrix <- dplyr::select(degs, c(paste0(metric, c('_PEF', '_REF', '_PEFREF'))))
row.names(expMatrix) <- degs$external_gene_name

#Generate a null model 
perms <- permute(as.data.frame(t(expMatrix)),  1000)$perm
nullModel <- as.data.frame(t(perms))

#Minimum number of target genes required to consider a regulon
minRegSize = 10

#Perform msVIPER for all three contrast using the desired metric (l2FC, z-score, or Wald statistic) A few parameters of relevance: 
# - minsize = minimum number of gene in a regulon for it to be included in the analysis
# - pleiotropy = if true, the NES for each regulon is penalized for a regulon's overlap with others

# Follow msVIPER with shawdow and leading edge anlysis 

resPEF <- msviper(as.matrix(expMatrix[paste0(metric, '_PEF')]), regulon, minsize = minRegSize, pleiotropy = TRUE, verbose = FALSE) %>%
  shadow() %>% ledge() 

resREF <-  msviper(as.matrix(expMatrix[paste0(metric, '_REF')]), regulon, minsize = minRegSize, pleiotropy = TRUE, verbose = FALSE) %>%
  shadow() %>% ledge() 

resPEFREF <-  msviper(as.matrix(expMatrix[paste0(metric, '_PEFREF')]), regulon, minsize = minRegSize, pleiotropy = TRUE, verbose = FALSE) %>%
  shadow() %>% ledge() 


#Build summary dataframes
resREF_sum <- data.frame('score' = resREF$es$nes, 'name' = names(resREF$es$nes), 'p_value' = resREF$es$p.value, 'p_adj' = p.adjust(resREF$es$p.value,method = 'fdr'))
resPEF_sum <- data.frame('score' = resPEF$es$nes, 'name' = names(resPEF$es$nes), 'p_value' = resPEF$es$p.value, 'p_adj' = p.adjust(resPEF$es$p.value, method = 'fdr'))
resPEFREF_sum <- data.frame('score' = resPEFREF$es$nes, 'name' = names(resPEFREF$es$nes), 'p_value' = resPEFREF$es$p.value, 'p_adj' = p.adjust(resPEFREF$es$p.value,method = 'fdr'))

#Build a summary table
resAll_sum <-  inner_join(resPEF_sum, resREF_sum, by = 'name', suffix = c('_PEF', '_REF')) %>% 
  inner_join(resPEFREF_sum, by = 'name') %>%
  dplyr::rename(score_PEFREF = score, p_value_PEFREF = p_value) %>%
  as.tibble() %>%
  add_column('PEF_cos_sim' = NA, 'REF_cos_sim' = NA, 'nes_composite_PEF' = NA, 'nes_composite_REF' = NA)


regMat <- filter(dorotheaPKN, tf %in% resAll_sum$name) %>%
  dplyr::rename(name = target) %>%
  inner_join(rownames_to_column(expMatrix, 'name'), by = 'name')

```

## Compute Cosine Composite Enrichment Score

To combine information between different contrasts and produce a composite enrichement score ($CES$), we will use the cosine similarity. Essentially we are accounting for how similar / different the enrichment of a given regulon is as opposed to adding the ES naively. 

$$
CES_{i} = \Bigg| \frac{\vec{r}_{(i, PvC)} \cdot  \vec{r}_{(i, PvR)}}{||\vec{r}_{(i, PvC)} || \text{ } || \vec{r}_{(i, PvR)}||} \Bigg| \bigg( \frac{NES_{(i, PvC)} + NES_{(i, PvR)}}{2}\bigg)
$$

```{r cosSimilarity}

for (i in (1:length(resAll_sum$name))){
  
  reg = resAll_sum$name[i]
  
  Ep = resAll_sum$score_PEF[i]
  Er = resAll_sum$score_REF[i]
  Epr = resAll_sum$score_PEFREF[i]
  
  pEF_id = paste0(metric, '_PEF')
  rEF_id = paste0(metric, '_REF')
  pEF_rEF_id = paste0(metric, '_PEFREF')
  
  cos_sim <- filter(regMat, tf == reg) %>%
    dplyr::select(pEF_id, rEF_id, pEF_rEF_id) %>%
    as.matrix() %>% 
    cosine() %>%
    abs()
  
  resAll_sum$PEF_cos_sim[i] = cos_sim[pEF_rEF_id, pEF_id]
  resAll_sum$REF_cos_sim[i] = -cos_sim[pEF_rEF_id, rEF_id]
  
  resAll_sum$nes_composite_PEF[i] <- cos_sim[pEF_rEF_id, pEF_id] * ((Ep) + (Epr)) / 2
  resAll_sum$nes_composite_REF[i]  <- cos_sim[pEF_rEF_id, rEF_id] * ((Er) - (Epr)) / 2
  
}

#Build unique metric dataframes
resPEF_nesc <- data.frame('score' =  resAll_sum$nes_composite_PEF, 'name' = resAll_sum$name, 'p_adj' = (resAll_sum$p_adj_PEF + resAll_sum$p_adj))
resREF_nesc <- data.frame('score' =  resAll_sum$nes_composite_REF, 'name' = resAll_sum$name)

```


## Results Plotting

```{r plottingTF_1}

# Make a helper function for the standard top TF plot

plotTopTFs <- function (res, topN = 20, title = NULL) {
    res <- arrange(res, desc(abs(score))) %>%
    dplyr::slice(1:topN)
  
   if (is.null(title)){
    title <- sprintf("Top %d Enriched Transcription Factors", topN)
   } else {
    title <- paste(sprintf("Top %d Enriched Transcription Factors -", topN), title)
  }
  
  ggplot(dplyr::select(res, c('name', 'score', 'p_adj'))) + 
      geom_bar(aes(x = reorder(name, (score)), y = score, fill = -log10(p_adj)), stat = "identity") +
    scale_fill_gradient2(name="-Log(Adj. p-value)", low = "darkblue", high = "steelblue1", 
        mid = "dodgerblue3", midpoint = (max(-log10(res$p_adj)) - min(-log10(res$p_adj))) / 2 + min(-log10(res$p_adj))) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size =10, face= "bold"),
    axis.text.y = element_text(size =10, face= "bold"),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()) +
    xlab(title) + 
    ylab('Enrichment Score')
}

#Plot for each contrast
plotTopTFs(resPEF_sum, title = 'HFpEF vs Control')
plotTopTFs(resREF_sum, title = 'HFrEF vs Control')
plotTopTFs(resPEFREF_sum, title = 'HFpEF vs HFrEF')

#Plot for "unique" metrics
plotTopTFs(resPEF_nesc, title = 'HFpEF - Cosine Composite NES')


setwd('~/Bioinformatics/Network Analysis of HFpEF/Hahn_FGN_Analysis/data')

#Update the names of the output, add z_scores
names(resAll_sum)[names(resAll_sum) == 'name'] <- 'external_gene_name'
resAll_sum$z_score_PEF = abs(qnorm(resAll_sum$p_value_PEF)) * sign(resAll_sum$score_PEF)
resAll_sum$z_score_REF = abs(qnorm(resAll_sum$p_value_REF)) * sign(resAll_sum$score_REF)
resAll_sum$z_score_PEFREF = abs(qnorm(resAll_sum$p_value_PEFREF)) * sign(resAll_sum$score_PEFREF)

#Save the results
write.csv(resAll_sum, paste0(metric, '_TF_enrichment.csv'), row.names = FALSE) 

```


## Samplewise Analysis

This allows us to extract correlations between TF enrichment and clinical characteristics. 

```{r sample_by_sample_VIPER}

vpsig <- viperSignature(exp_dset, 'disease', 'Normal', per = 1000)

vpres <- viper(vpsig, regulon, verbose = TRUE, minsize = 10)

pos <- pData(vpres)[["disease"]] %in% c("HFpEF", "HFrEF", "Normal")
res_viper <- t(exprs(vpres)[, pos]) %>%
  as.data.frame() 

#Pickup the clinical correlates to compare to TF enrichment
res_viper['Disease'] <- pData(vpres)[["disease"]]
res_viper['BMI'] <- pData(vpres)[["bmi"]] 
res_viper['T2DM'] <- pData(vpres)[["diabetes"]]
res_viper['Age'] <- pData(vpres)[["age"]]
res_viper['Sex'] <- pData(vpres)[["sex"]]
res_viper['eGFR'] <- pData(vpres)[["eGFR"]]

# Save results
setwd('~/Bioinformatics/Network Analysis of HFpEF/Hahn_FGN_Analysis/data')
write.csv(res_viper, paste0(metric, '_samplewise_TF_enrichment.csv'), row.names = FALSE) 
```
## IRF3 Activity Correlations with Clinical Characteristics 

For IRF3 specifically, compute correlations for both the enrichment score and exression with clinical covariates

IRF3 is the top enriched TF for HFpEF using the cosine composite enrichment score. Let's examine this a bit closer by looking for correlations against given clinical covariates. 

```{r correlations}


# Patient difference heat map
d1 <- exprs(vpres)[, pos]
colnames(d1) <- pData(vpres)[["Disease"]][pos]
dd <- dist(t(d1), method = "euclidean")
heatmap(as.matrix(dd), Rowv = as.dendrogram(hclust(dd, method = "average")), symm = T)

myColors <- c("#ffb3ba", "#ffffba")


# IRF3 - TF enrichment correlations
ggplot(res_viper, aes(x=BMI, y=IRF3)) +
  geom_point(aes(color = Disease, shape = Sex), size = 3.25) + 
  geom_smooth(method= lm, fullrange = TRUE)

ggpairs(res_viper, columns = c('IRF3', 'BMI', 'eGFR', 'Age'), alpha = 1, ggplot2::aes(colour=Disease, alpha = 0.9)) 

#IRF3 - expression correlations
cnts_plot <- t(exp_dset@assayData$exprs) %>%
  as.data.frame() %>%
  dplyr::select(IRF3)
cnts_plot['Disease'] <- exp_dset$disease
cnts_plot['BMI'] <- exp_dset$bmi
cnts_plot['T2DM'] <- exp_dset$diabetes
cnts_plot['Age'] <- exp_dset$age
cnts_plot['Sex'] <- exp_dset$sex
cnts_plot['eGFR'] <- exp_dset$eGFR

ggplot(cnts_plot, aes(x=BMI, y=IRF3)) +
  geom_point(aes(color = Disease, shape = Sex), size = 3.25) + 
  geom_smooth(method= lm, fullrange = TRUE)

ggpairs(cnts_plot, columns = c('BMI', 'eGFR', 'Age','IRF3'), alpha = 1, ggplot2::aes(colour=Disease, alpha = 0.9)) 

```

